\cfoot{}\rhead{\thepage}
\lhead{{\scshape Solovay's model} $\qquad$ {\tiny \today } }

% \noindent{\Large{\scshape\bfseries Entreé to Generic Extensions and Forcing in Set Theory}} \\[0.1cm]
%
% \noindent {\scshape Bohuslav Balcar}, {\small CTS, J{\' \i}lsk{\' a} 1, Praha 1,
% 	Czech Republic, {\ttfamily balcar@cts.cuni.cz} } \\[0.1cm]
% \noindent {\scshape Jonathan Verner}, {\small KTIML MFF UK, {\ttfamily jonathan.verner@matfyz.cz}\\[0.1cm]
% {\tiny \today } \\[0.5cm]

%\maketitle

\thispagestyle{empty}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Solovay's model, where all sets are measurable}\label{solovay_chapter}

{\tiny \today } \\[0.5cm]

Every matematicien met the Lebesgue measure and integral and is
familier with several construction of non-measurable sets, eg.
Vitali's construction, Bernstein sets, nontrivial ultrafilter
on $\omega$ and is aware that construction of such a set is
not effective and some form of axiom of choice is needed.

On the other hand if one is trying to show some desired properties
of measure such that the countable union of null sets is a null set,
i.e. that the ideal of null sets is $\sigma$-ideal, some form
of the axiom of choice is also needed. To show this property
it is enough to assume DC, or even weaker version DCR (dependent
choice for subsets of reals).

R.~Solovay in \cite{Sol:1970} constructed ZF model $M$ in which
DC holds and all subsets of real line are Lebesgue measurable.
Initial assumption to Solovay construction is the existence of
inaccessible cardinal number $\kappa$. The starting point is
generic extension of such ZFC model using L{\' e}vy collaps
Levi$(\omega,<\kappa)$ forcing to  obtain model $N = V[G]$.
Then the desired model is obtained using different methods
as a inner model of $N$, this model is called a Solovay model.

Later, S.~Shelah and Raisiniere showed that the assumption
of an inaccessible cardinal is necessary. In fact from the existence of a model
where all $\Sigma_3^1$ sets from projective hierarchy are
measurable follows the existence of an inaccessible cardinal
in some inner model. Shortly, the fact that all sets are measurable
is equiconsistent with the existence of an inaccessible cardinal.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%                          END                                    %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%